---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# parafac4microbiome

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/GRvanderPloeg/parafac4microbiome/branch/master/graph/badge.svg)](https://app.codecov.io/gh/GRvanderPloeg/parafac4microbiome?branch=master)
[![R-CMD-check](https://github.com/GRvanderPloeg/parafac4microbiome/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/GRvanderPloeg/parafac4microbiome/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Overview
The parafac4microbiome package enables R users with an easy way to create Parallel Factor Analysis (PARAFAC) models for longitudinal microbiome data.

*    `processDataCube()` can be used to process the microbiome count data appropriately for a multi-way data array.
*    `parafac()` allows the user to create a Parallel Factor Analysis model of the multi-way data array.
*    `assessNumComponents()` helps the user select the appropriate number of components for the PARAFAC model.
*    `modelStabilityCheck()` performs jack-knifing of samples to inspect the stability of the PARAFAC model.
*    `plotPARAFACmodel()` helps visually inspect the PARAFAC model.

This package also comes with three example datasets.

*    `Fujita2023`: an in-vitro experiment of ocean inocula on peptide medium, sampled every day for 110 days (https://doi.org/10.1186/s40168-023-01474-5).
*    `Shao2019`: a large cohort dataset of vaginally and caesarean-section born infants from London (https://www.nature.com/articles/s41586-019-1560-1).
*    `vanderPloeg2024`: a small gingivitis intervention dataset with specific response groups (manuscript in preparation).

A basic introduction to the package is given in `vignette("PARAFAC_introduction")` and modelling the example datasets are elaborated in their respective vignettes `vignette("Fujita2023_analysis")`, `vignette("Shao2019_analysis")` and `vignette("vanderPloeg2024_analysis")`.

## Installation

You can install the development version of parafac4microbiome from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("GRvanderPloeg/parafac4microbiome")
```
## Citation
Please use the following citation when using this package:

*   van der Ploeg, G. R., et al. (2024). Multi-way modelling of oral microbial dynamics and host-microbiome interactions during induced gingivitis. [Manuscript in preparation].

## Usage

```{r usage}
library(parafac4microbiome)
library(multiway)
library(dplyr)
library(ggplot2)
library(ggpubr)
set.seed(0) # for reproducibility

# Process the data cube
processedFujita = processDataCube(Fujita2023, sparsityThreshold=0.99, centerMode=1, scaleMode=2)

# Make a PARAFAC model
model = parafac(processedFujita$data, nfac=3, verbose=FALSE)

# Plot the PARAFAC model using some metadata
plotPARAFACmodel(model, processedFujita,
  colourCols = c("", "Genus", ""),
  legendTitles = c("", "Genus", ""),
  xLabels = c("Replicate", "Feature index", "Time point"),
  legendColNums = c(0,5,0),
  arrangeModes = c(FALSE, TRUE, FALSE),
  continuousModes = c(FALSE,FALSE,TRUE),
  overallTitle = "Fujita PARAFAC model")
```
