---
title: "Fujita2023 analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fujita2023_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 6
)
options(tibble.print_min = 6L, tibble.print.max = 6L, digits = 3)
```

# Introduction
In this vignette we examine and model the Fujita2023 data in more detail.

```{r setup}
library(parafac4microbiome)
library(dplyr)
library(ggplot2)
library(ggpubr)
```

# Processing the data cube
The data cube in Fujita2023$data contains unprocessed counts. The function processDataCube() performs the processing of these counts with the following steps:

*   It performs feature selection based on the sparsityThreshold setting. Sparsity is here defined as the fraction of samples where a microbial abundance (ASV/OTU or otherwise) is zero.
*   It performs a centered log-ratio transformation of each sample using the compositions::clr() function with a pseudo-count of one (on all features, prior to selection based on sparsity).
*   It centers and scales the three-way array. This is a complex subject, for which we refer to a [paper by Rasmus Bro and Age Smilde](https://doi.org/10.1002/cem.773). By centering across the subject mode, we make the subjects comparable to each other within each time point. Scaling within the feature mode avoids the PARAFAC model focussing on features with abnormally high variation.

The outcome of processing is a new version of the dataset. Please refer to the documentation of processDataCube() for more information.

```{r data processing}
processedFujita = processDataCube(Fujita2023, sparsityThreshold=0.99, centerMode=1, scaleMode=2)
```

# Determining the correct number of components
A critical aspect of PARAFAC modelling is to determine the correct number of components. We have developed the function assessNumComponents() for this purpose. We give the function our new dataset and specify the minimum and maximum number of components to investigate and the number of randomly initialized models to try for each number of components.

Note: this vignette reflects a minimum working example for analysing this dataset due to computational limitations. Hence, we only look at 1-3 components with 5 random initializations each. These settings are not ideal for real datasets. Please refer to the documentation of assessNumComponents() for more information.

```{r Fujita num comp selection}
# Setup
minNumComponents = 1
maxNumComponents = 3
numRepetitions = 5 # number of randomly initialized models
numFolds = 8 # number of jack-knifed models
ctol = 1e-6
maxit = 200
numCores= 1

# Plot settings
colourCols = c("", "Genus", "")
legendTitles = c("", "Genus", "")
xLabels = c("Replicate", "Feature index", "Time point")
legendColNums = c(0,5,0)
arrangeModes = c(FALSE, TRUE, FALSE)
continuousModes = c(FALSE,FALSE,TRUE)

# Assess the metrics to determine the correct number of components
assessment = assessNumComponents(processedFujita$data, minNumComponents, maxNumComponents, numRepetitions, ctol=ctol, maxit=maxit, numCores=numCores)
```

We will now inspect the output plots of interest for Fujita2023.

```{r overview plot}
assessment$plots$overview
```
The overview plots show that we can reach ~35% explained variation if we take 3 components. The CORCONDIA for those models are ~98.5, which is well above the minimum requirement of 60. Based on this overivew, either 2 or 3 components seems fine.

```{r TCC}
assessment$plots$TCC[[2]]
assessment$plots$TCCoverall[[2]]
assessment$plots$TCC[[3]]
assessment$plots$TCCoverall[[3]]
```

The Tucker Congruence Coefficients show us how similar the loading vectors in the same mode are to each other, between the various components. A value >=0.85 has been described as problematic in the literature. Neither in the two-component model, nor in the three-component model are the loadings overly similar to each other.

# Jack-knifed models
Next, we investigate the stability of the models when jack-knifing out samples. This will give us more information to choose between 2 or 3 components.

```{r model stability}
modelStabilityCheck(processedFujita, numComponents=2, numFolds=numFolds, considerGroups=FALSE,
                    groupVariable="", colourCols, legendTitles, xLabels, legendColNums, arrangeModes,
                    continuousModes, ctol=ctol, maxit=maxit, numCores=numCores)$plot
modelStabilityCheck(processedFujita, numComponents=3, numFolds=numFolds, considerGroups=FALSE,
                    groupVariable="", colourCols, legendTitles, xLabels, legendColNums, arrangeModes,
                    continuousModes, ctol=ctol, maxit=maxit, numCores=numCores)$plot
```
In the two-component jack-knifed models, there seems to be some instability in the second component, mainly in the time mode. In the three-component models there is more stability overall.

# Model selection
We have decided that a three-component model is the most appropriate for the Fujita2023 dataset. We can now select one of the random initializations from the assessNumComponents() output as our final model. We're going to select the random initialisation that corresponded the maximum amount of variation explained for three components.
```{r model selection}
numComponents = 3
modelChoice = which(assessment$metrics$varExp[,numComponents] == max(assessment$metrics$varExp[,numComponents]))
finalModel = assessment$models[[numComponents]][[modelChoice]]
```

Finally, we visualize the model using plotPARAFACmodel().

```{r model plot}
plotPARAFACmodel(finalModel, processedFujita, colourCols, legendTitles, xLabels, legendColNums, arrangeModes,
  continuousModes = c(FALSE,FALSE,TRUE),
  overallTitle = "Fujita PARAFAC model")
```
