---
title: "MAINHEALTH: intergenerational obesity transmission"
output: html_document
date: "2025-08-31"
---

# Introduction

This article presents the CP results of the MAINHEALTH dataset discussed in Chapter 5 of my dissertation.

# Preamble

```{r setup}
library(dplyr)
library(tidyr)
library(parafac4microbiome)
library(NPLStoolbox)
library(CMTFtoolbox)
library(ggplot2)
library(ggpubr)
library(stringr)
library(scales)
library(ggpattern)
```

```{r metadata test function}
testMetadata = function(model, comp, metadata){
  transformedSubjectLoadings = model$Fac[[1]][,comp]
  transformedSubjectLoadings = transformedSubjectLoadings / norm(transformedSubjectLoadings, "2")
  
  result = lm(transformedSubjectLoadings ~ BMI + C.section + Secretor + Lewis + whz.6m, data=metadata$mode1)
  
  # Extract coefficients and confidence intervals
  coef_estimates <- summary(result)$coefficients
  conf_intervals <- confint(result)
  
  # Remove intercept
  coef_estimates <- coef_estimates[rownames(coef_estimates) != "(Intercept)", ]
  conf_intervals <- conf_intervals[rownames(conf_intervals) != "(Intercept)", ]
  
  # Combine into a clean data frame
  summary_table <- data.frame(
    Term     = rownames(coef_estimates),
    Estimate = coef_estimates[, "Estimate"] * 1e3,
    CI       = paste0(
                 conf_intervals[, 1], " – ",
                 conf_intervals[, 2]
               ),
    P_value  = coef_estimates[, "Pr(>|t|)"],
    P_adjust = p.adjust(coef_estimates[, "Pr(>|t|)"], "BH"),
    row.names = NULL
  )
  
  return(summary_table)
}
```

```{r define colours}
colours = RColorBrewer:: brewer.pal(8, "Dark2")
BMI_cols = c("darkgreen","darkgoldenrod","darkred") #colours[1:3]
WHZ_cols = c("darkgreen", "darkgoldenrod", "dodgerblue4")
phylum_cols = colours
```

```{r process data}
processedFaeces = parafac4microbiome::processDataCube(NPLStoolbox::Jakobsen2025$faeces, sparsityThreshold = 0.75, centerMode=1, scaleMode=2)
processedMilk = parafac4microbiome::processDataCube(NPLStoolbox::Jakobsen2025$milkMicrobiome, sparsityThreshold=0.85, centerMode=1, scaleMode=2)
processedMilkMetab = parafac4microbiome::processDataCube(NPLStoolbox::Jakobsen2025$milkMetabolomics, CLR=FALSE, centerMode=1, scaleMode=2)
```

```{r plotting variables}
colourCols = c("", "", "")
legendTitles = c("BMI group", "Phylum", "")
xLabels = c("Subject index", "Feature index", "Time index")
legendColNums = c(3,5,0)
arrangeModes = c(TRUE, TRUE, FALSE)
continuousModes = c(FALSE,FALSE,TRUE)
```

# Faecal microbiome

CP was applied to capture the dominant source of variation in the infant faecal microbiome data block. The model selection procedure revealed that CORCONDIA was inconclusive and loadings were somewhat unstable at three components. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r faecal model selection}
# assessment_faeces = parafac4microbiome::assessModelQuality(processedFaeces$data, numRepetitions = 10, numCores = 10)
# assessment_faeces$metrics$CORCONDIA %>% as_tibble() %>% mutate(index=1:nrow(.)) %>% pivot_longer(-index) %>% ggplot(aes(x=as.factor(name),y=value)) + geom_boxplot() + xlab("Number of components") + ylab("CORCONDIA") + theme(text=element_text(size=16))

# stability_faeces = parafac4microbiome::assessModelStability(processedFaeces, maxNumComponents=3, numFolds=20, colourCols=colourCols, legendTitles=legendTitles, xLabels = xLabels, legendColNums=legendColNums, arrangeModes=arrangeModes, numCores=parallel::detectCores())
# stability_faeces$modelPlots[[2]]
# stability_faeces$modelPlots[[3]]
# 

```

```{r faecal load model}
# cp_faeces = parafac4microbiome::parafac(processedFaeces$data, nfac=3, nstart = 100)
cp_faeces = readRDS("./MAINHEALTH/cp_faeces.RDS")

cp_faeces$varExp
```

Therefore, a two-component CP model was selected, explaining 15.2% of the variation. However, MLR analysis revealed that neither component could be interpreted.

```{r faecal cp subject mode metadata test}
testMetadata(cp_faeces, 1, processedFaeces)
testMetadata(cp_faeces, 2, processedFaeces)
```

# HM microbiome

CP was applied to capture the dominant source of variation in the HM microbiome data block. The model selection procedure revealed that CORCONDIA was inconclusive and loadings were unstable at three components. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r hm microbiome model selection}
# assessment_milk = parafac4microbiome::assessModelQuality(processedMilk$data, numRepetitions = 10, numCores = 10)
# assessment_milk$metrics$CORCONDIA %>% as_tibble() %>% mutate(index=1:nrow(.)) %>% pivot_longer(-index) %>% ggplot(aes(x=as.factor(name),y=value)) + geom_boxplot() + xlab("Number of components") + ylab("CORCONDIA") + theme(text=element_text(size=16))

# stability_milk = parafac4microbiome::assessModelStability(processedMilk, maxNumComponents=3, numFolds=10, colourCols=colourCols, legendTitles=legendTitles, xLabels = xLabels, legendColNums=legendColNums, arrangeModes=arrangeModes, numCores=parallel::detectCores())
# stability_milk$modelPlots[[2]]
# stability_milk$modelPlots[[3]]
# 

```

```{r hm microbiome load model}
# cp_milk = parafac4microbiome::parafac(processedMilk$data, nfac=2, nstart = 100)
cp_milk = readRDS("./MAINHEALTH/cp_milk.RDS")

cp_milk$varExp
```

Therefore, a two-component CP model was selected, explaining 13.2% of the variation. However, MLR analysis revealed that the first CP component was uninterpretable, while the second captured variation only weakly associated with maternal ppBMI.

```{r hm microbiome cp subject mode metadata test}
testMetadata(cp_milk, 1, processedMilk)
testMetadata(cp_milk, 2, processedMilk)
```

```{r cp HM milk comp 2}
topIndices = processedMilk$mode2 %>% mutate(index=1:nrow(.), Comp = cp_milk$Fac[[2]][,2]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
bottomIndices = processedMilk$mode2 %>% mutate(index=1:nrow(.), Comp = cp_milk$Fac[[2]][,2]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()

Xhat = parafac4microbiome::reinflateTensor(cp_milk$Fac[[1]][,2], cp_milk$Fac[[2]][,2], cp_milk$Fac[[3]][,2])

print("Positive loadings:")
print(cor(Xhat[,topIndices[1],1], processedMilk$mode1$BMI)) # no flip
print("Negative loadings:")
print(cor(Xhat[,bottomIndices[1],1], processedMilk$mode1$BMI))

a = processedMilk$mode1 %>%
  mutate(Component_2 = cp_milk$Fac[[1]][,2]) %>%
  ggplot(aes(x=BMI,y=Component_2)) +
  geom_point() +
  stat_cor() +
  xlab("ppBMI") +
  ylab("Loading") +
  theme(text=element_text(size=16))

df = processedMilk$mode2 %>%
  mutate(Component_2 = cp_milk$Fac[[2]][,2]) %>%
  arrange(Component_2) %>%
  mutate(index=1:nrow(.)) %>%
  mutate(Genus = gsub("g__", "", V7), Species = gsub("s__", "", V8)) %>% 
  mutate(Species = str_split_fixed(Species, "_", 2)[,2]) %>% 
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "sp."), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "organism"), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "bacterium"), "", .x))) %>%
  filter(index %in% c(1:10, 106:115))

df[df$Genus == "" & df$Species == "", "Species"] = "Unknown"
df[df$Species == "", "Species"] = "sp."
df$name = paste0(df$Genus, " ", df$Species)
  
b = df %>% ggplot(aes(x=Component_2,y=as.factor(index),fill=as.factor(V3))) +
  geom_bar(stat="identity", col="black") +
  xlab("Loading") +
  ylab("") +
  scale_y_discrete(labels=df$name) +
  scale_fill_manual(name="Phylum", values=hue_pal()(5)[-c(2,5)], labels=c("Actinobacteriota", "Firmicutes", "Proteobacteria")) +
  theme(text=element_text(size=16))

c = cp_milk$Fac[[3]][,2] %>%
  as_tibble() %>% 
  mutate(value=value) %>% 
  ggplot(aes(x=c(0,30,60,90),y=value)) +
  geom_line() +
  geom_point() +
  xlab("Time point [day]") +
  ylab("Loading") +
  ylim(0,1) +
  theme(text=element_text(size=16))

a
b
c
```

# HM metabolome

CP was applied to capture the dominant source of variation in the HM metabolomics data block. The model selection procedure revealed that CORCONDIA values were too low at four components. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r hm metabolome model selection}
# assessment_milkMetab = parafac4microbiome::assessModelQuality(processedMilkMetab$data, numRepetitions = 10, numCores = 10)
# 
# assessment_milkMetab$metrics$CORCONDIA %>% as_tibble() %>% mutate(index=1:nrow(.)) %>% pivot_longer(-index) %>% ggplot(aes(x=as.factor(name),y=value)) + geom_boxplot() + xlab("Number of components") + ylab("CORCONDIA") + theme(text=element_text(size=16))

# stability_milkMetab = parafac4microbiome::assessModelStability(processedMilkMetab, maxNumComponents=4, numFolds=10, colourCols=c("", "", ""), legendTitles=c("BMI group", "Class", ""), xLabels = xLabels, legendColNums=legendColNums, arrangeModes=arrangeModes, numCores=parallel::detectCores())
# stability_milkMetab$modelPlots[[3]]
# stability_milkMetab$modelPlots[[4]]
```

```{r hm metabolome load model}
# cp_milkMetab = parafac4microbiome::parafac(processedMilkMetab$data, nfac=3, nstart = 100)
cp_milkMetab = readRDS("./MAINHEALTH/cp_milkMetab.RDS")

cp_milkMetab$varExp
```

Therefore, a three-component CP model was selected, explaining 21.6% of the variation. MLR analysis revealed that the first and third components were uninterpretable, while the second component captured variation exclusively and strongly associated with maternal secretor status.

```{r hm metabolome cp subject mode metadata test}
testMetadata(cp_milkMetab, 1, processedMilkMetab)
testMetadata(cp_milkMetab, 2, processedMilkMetab)
testMetadata(cp_milkMetab, 3, processedMilkMetab)
```

```{r cp milkMetab comp 2}
topIndices = processedMilkMetab$mode2 %>% mutate(index=1:nrow(.), Comp = cp_milkMetab$Fac[[2]][,2]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
bottomIndices = processedMilkMetab$mode2 %>% mutate(index=1:nrow(.), Comp = cp_milkMetab$Fac[[2]][,2]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()

Xhat = parafac4microbiome::reinflateTensor(cp_milkMetab$Fac[[1]][,2], cp_milkMetab$Fac[[2]][,2], cp_milkMetab$Fac[[3]][,2])

print("Positive loadings:")
print(cor(Xhat[,topIndices[1],1], processedMilkMetab$mode1$Secretor)) # flip
print("Negative loadings:")
print(cor(Xhat[,bottomIndices[1],1], processedMilkMetab$mode1$Secretor))

a = processedMilkMetab$mode1 %>%
  mutate(Component_2 = cp_milkMetab$Fac[[1]][,2]) %>%
  ggplot(aes(x=as.factor(Secretor),y=Component_2)) +
  geom_boxplot() +
  stat_compare_means() +
  xlab("Secretor status") +
  ylab("Loading") +
  scale_x_discrete(labels=c("Se-", "Se+")) +
  theme(text=element_text(size=16))

temp = processedMilkMetab$mode2 %>%
  mutate(Component_2 = -1*cp_milkMetab$Fac[[2]][,2]) %>%
  arrange(Component_2) %>%
  mutate(index=1:nrow(.)) %>%
  filter(index %in% c(1:10, 61:70))

b=temp %>%
  ggplot(aes(x=Component_2,y=as.factor(index),fill=as.factor(Class),pattern=as.factor(Class))) +
  geom_bar_pattern(stat="identity",
                   colour="black",
                   pattern="stripe",
                   pattern_fill="black",
                   pattern_density=0.2,
                   pattern_spacing=0.05,
                   pattern_angle=45,
                   pattern_size=0.2) +
  scale_y_discrete(label=temp$Metabolite) +
  xlab("Loading") +
  ylab("") +
  scale_fill_manual(name="Class", values=hue_pal()(7)[-c(4,5)], labels=c("Amino acids and derivatives", "Energy related", "Fatty acids and derivatives", "Oligosaccharides", "Sugars")) +
  theme(text=element_text(size=16))

c = cp_milkMetab$Fac[[3]][,2] %>%
  as_tibble() %>% 
  mutate(value=value*-1) %>% 
  ggplot(aes(x=c(0,30,60,90),y=value)) +
  geom_line() +
  geom_point() +
  xlab("Time point [day]") +
  ylab("Loading") +
  ylim(0,1) +
  theme(text=element_text(size=16))

a
b
c
```
